Adaptive group-regularized logistic elastic net regression Supplementary material

### 2019-03-01 docs/manuscript.Rnw
ran micrornaseq_colorectal_cancer.R 10 random groups and stability selection

	An example of such a study with extra information available is investigated in
	Sections [! section refs here]. The first one is concerned with treatment 
	response prediction for colorectal cancer patients, based on sequenced
	microRNAs. In the second example we attempt to predict lymph node metastases,
	using RNAseq data. In both settings we have additional information on the 
	features in the form of a partitioning available. Furthermore, in both
	examples data is sparse, so we would like to 
	include the additional information to enhance predictive performance of our
	models. Inclusion of the partitions is not straightforward using standard 
	methods available in literature. At first sight, such
	an \textit{a priori} grouping of the features suggests the group lasso
	\cite[]{meier_group_2008, yuan_model_2006} (or one of its extensions), which
	penalises and selects features at the group level. There are two limitations 
	to the group lasso: (a) it selects groups as a whole instead of single 
	features and (b) it is not adaptive: the amount of penalization a group
	receives depends on a `global' penalty parameter common to all groups.
	Extensions such as the sparse group lasso \cite[]{simon_sparse-group_2013}
	partly deal with (a), but do not address (b).

@@

	\section{Application to microRNAs in colorectal cancer}\label{sec:appcolon}
  The presented data is from a deep sequencing analysis on microRNAs 
  \cite[]{neerincx_combination_2018}. The study was done in 88 treated 
  colorectal cancer patients, with the aim of predicting treatment response,
  coded as either non-progressive/remission (70 patients) or progressive 
  (18 patients). After pre-processing and normalisation, 2114 microRNAs 
  remained. In addition to the 2114 microRNAs, four clinical covariates were
  included in the analysis: prior use of adjuvant therapy (binary),
  the type of systemic treatment regimen (ternary), age, and primary tumor
  differentiation (binary). These clinical covariates are included as 
  unpenalized features in the model.

  In a preliminary experiment on different subjects, the microRNA expression 
  levels of metastatic colorectal tumour tissue were compared to normal 
  non-colorectal tissue and primary colorectal tumour tissue was compared to
  primary colorectal tumour tissue \cite[]{neerincx_mir_2015}. This yielded 221
  microRNAs that were differentially expressed in both comparisons ($\text{FDR}
  \leq 0.05$), versus 1893 not differentially expressed microRNAs. We expect
  that incorporation of this partitioning enhances therapy response 
  classification, because tumor-specific miRNAs might be more relevant than 
  non-specific ones. In addition, we divided the differentially expressed
  microRNAs even further into 127 highly differentially expressed microRNAs
  ($\text{FDR} \leq 0.001$) and 94 medium differentially expressed microRNAs 
  ($0.001 < \text{FDR} \leq 0.05$).  We will refer to this second partitioning
  as the three-group setting, as opposed to the first, the two-group setting.

  We split the data into 61 training, and 27 test instances and
  compared the performance of \texttt{gren} on the test set to ridge, 
  \texttt{GRridge}, elastic net, and the composite minimax
  concave penalty (\texttt{cMCP}) by \cite{breheny_penalized_2009}. Further
  comparisons to the sparse
  group lasso (\texttt{SGL}) by \cite{simon_sparse-group_2013} and the group
  exponential lasso (\texttt{gel}) \cite[]{breheny_group_2015} are presented in 
  the SM [! section ref here]. We compared the methods on AUC. More performance
  measures are presented in the SM [! section ref here].
  For the elastic net methods (including \texttt{gren}) and \texttt{cMCP}, we 
  set $\alpha \in \{ 0.05, 0.5, 0.95\}$. We experienced numerical issues with 
  \texttt{GRridge} when cross-validating 
  the global penalty parameter on the training data. We therefore estimated this
  global parameter on the full data for \texttt{GRridge}.

  The results are presented in Figure 
  \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res2_auc}. 
  The penalty multipliers estimated by 
  \texttt{gren} and \texttt{GRridge} are according to expectation: the highly
  expressed group receives the lowest penalty, followed by the medium expressed
  group, while the non-expressed group receives the strongest penalty. 
  \texttt{gren} seems to outperform the other methods for a range of model 
  sizes. This is especially pronounced for the larger models. The performance of
  \texttt{cmcp} is somewhat unstable. We noticed this is in general
  for the group-lasso and its extensions (see SM [! section ref here]). Overall,
  it seems as though the inclusion of the extra information on the features 
  benefited predictive performance: both \texttt{GRridge} and \texttt{gren} 
  outperform their respective non-group-regularized versions.
<<barplot_lines_micrornaseq_colorectal_cancer_res2_auc, fig.cap="Results for the colorectal cancer example with (a) the estimated penalty multipliers and (b) estimated AUC.", out.width="100%", fig.asp=1/2>>=
@

  \section{Application to RNAseq in lymph node cancer}\label{sec:applymph}
  In this application, the aim is to predict lymph node metastasis in patients 
  with HPV negative oral cancer using sequenced RNAs from TCGA 
  \cite[]{the_cancer_genome_atlas_network_comprehensive_2015}. The features are 
  normalised TCGA RNASeqv2 profiles of head-and-neck squamous cell 
  carcinomas on 133 patients with HPV-negative tumours in the oral cavity. Of 
  these patients, 76 had lymph node metastasis, and 57 did not. For a thorough 
  introduction of these data, see \cite{te_beest_improved_2017}.

  As co-data we considered two sources: (a) the cis-correlation between TCGA DNA
  copy number measurements and the RNASeqv2 data on the same patients, 
  quantified through Kendall's $\tau$ and binned into five groups, and (b) 
  p-values from an external microarray data set \cite[]{mes_prognostic_2017}, 
  binned into five groups. We expect features 
  with a positive Kendall's $\tau$ to be more important in metastasis prediction   
  \cite[]{masayesva_gene_2004}. Likewise, we expect features with a low 
  p-value to be more important.

  We split our data into 120 training instances and 97 test instances, trained
  the models on the training set and estimated AUC on the test set. Other 
  performance measures are presented in the SM Section [! section ref here]. We
  compared the same models as in Section \ref{sec:appcolon}. In this 
  application, the co-data consists of two partitions with overlapping groups.
  \texttt{cMCP} is not designed to deal with overlapping groups, so we 
  cross-tabulate the two co-data partitions to create one paritioning with 25 
  groups for \texttt{cMCP}. In addition, we compare to an extension of 
  \texttt{cMCP} that is able to handle multiple groupings 
  \cite[]{zeng_overlapping_2016}. We present the results in Figure 
  \ref{fig:barplot_lines_rnaseq_lymph_node_metastasis_res1_auc}. 
  Additional comparisons to \texttt{SGL}, \texttt{gel} and 
  the overlapping group version of \texttt{gel} \cite[]{zeng_overlapping_2016}
  \cite[]{breheny_group_2015} are found in the SM Section [! section ref here].
<<barplot_lines_rnaseq_lymph_node_metastasis_res1_auc, fig.cap="Results for the lymph node metastasis example with (a) the estimated penalty multipliers for the cis-correlations, (b) the estimated penalty multipliers for the p-values, and (c) the estimated AUC.", out.width="100%", fig.asp=1>>=
@

  We present additional applications and comparisons in the SM Sections 
  [! sections ref here].

@@

	Although our method can be considered weakly adaptive, it is different from 
	the adaptive lasso \cite[]{van_de_geer_adaptive_2011,huang_adaptive_2008,
	zhang_adaptive_2007,zou_adaptive_2006} and adaptive elastic net 
	\cite[]{zou_adaptive_2009} in the sense that it adapts to external information
	rather than to the data. It also differs in the scale of adaptation: in the 
	adaptive lasso and elastic net the adaptive weights are feature specific,
	while in our case they are estimated on the group level, rendering the 
	adaptation more robust. As can be seen from the simulations and real data 
	example in Sections \ref{sec:simulations} and [! section ref here], 
	adaptation to 
	external data may be beneficial for prediction and feature selection. We 
	believe that this is due to the `borrowing of information' effect: estimates 
	that are believed to be similar are shrunken in the same way, yielding 
	overall, better estimates.

@@

	\section*{Software}
	A (stable) \texttt{R} package is available from 
	\url{https://CRAN.R-project.org/package=gren}.

	\section*{Supplementary Material}
	Supplementary Material is available online from 
	\url{http://pub.math.leidenuniv.nl/~munchmm/}. All materials needed to   
	recreate this document are available from 
	\url{https://github.com/magnusmunch/gren}.

	\section*{Acknowledgements}
	\textit{Conflict of Interest}: None declared.

>>>>

	Examples of such a study with extra information available are investigated in
	Sections [! section refs here]. The first one is concerned with treatment 
	response prediction for colorectal cancer patients, based on sequenced
	microRNAs. In the second example we attempt to predict lymph node metastases,
	using RNAseq data. In both settings we have additional information on the 
	features in the form of a partitioning available. Furthermore, in both
	examples data is sparse, so we would like to 
	include the additional information to enhance predictive performance of our
	models. Inclusion of the partitions is not straightforward using standard 
	methods available in literature. At first sight, such
	an \textit{a priori} grouping of the features suggests the group lasso
	\cite[]{meier_group_2008, yuan_model_2006} (or one of its extensions), which
	penalises and selects features at the group level. There are two limitations 
	to the group lasso: (a) it selects groups as a whole instead of single 
	features and (b) it is not adaptive: the amount of penalization a group
	receives depends on a `global' penalty parameter common to all groups.
	Extensions such as the sparse group lasso \cite[]{simon_sparse-group_2013}
	partly deal with (a), but do not address (b).

@@

	\section{Application to microRNAs in colorectal cancer}\label{sec:appcolon}
  The presented data is from a deep sequencing analysis on microRNAs 
  \cite[]{neerincx_combination_2018}. The study was done in 88 treated 
  colorectal cancer patients, with the aim of predicting treatment response,
  coded as either non-progressive/remission (70 patients) or progressive 
  (18 patients). After pre-processing and normalisation, 2114 microRNAs 
  remained. In addition to the 2114 microRNAs, four clinical unpenalized 
  covariates were included in the analysis: prior use of adjuvant therapy 
  (binary), the type of systemic treatment regimen (ternary), age, and primary 
  tumor differentiation (binary).

  In a preliminary experiment on different subjects, the microRNA expression 
  levels of metastatic colorectal tumour tissue were compared to normal 
  non-colorectal tissue and primary colorectal tumour tissue was compared to
  primary colorectal tumour tissue \cite[]{neerincx_mir_2015}. This yielded 127
  microRNAs that were differentially expressed in both comparisons with
  $\text{FDR} \leq 0.001$, 95 with and $\text{FDR} \leq 0.05$, and 1893 not 
  differentially expressed microRNAs. We expect that incorporation of this 
  partitioning enhances therapy response 
  classification, because tumor-specific miRNAs might be more relevant than 
  non-specific ones.

  We compared the performance of \texttt{gren} to ridge, \texttt{GRridge}, 
  elastic net, and to the best performing
  group lasso extension, the composite minimax
  concave penalty (\texttt{cMCP}) by \cite{breheny_penalized_2009}. 
  For the methods with an extra tuning parameter 
  $\alpha$, we picked the one (or two) that gave the best performance and refer
  the reader to SM Section [! section ref here] for more comparisons to 
  different $\alpha$s and to the sparse
  group lasso (\texttt{SGL}) by \cite{simon_sparse-group_2013} and the group
  exponential lasso (\texttt{gel}) \cite[]{breheny_group_2015}.

  To estimate performance, we split the data into 61 training, and 27 test 
  instances stratified by treatment response. We estimated the models on the 
  training data and calculated AUC on the test data. We present AUC for a range 
  of model sizes, together with the estimated 
  penalty multipliers for \texttt{gren} and \texttt{GRridge} in Figure
  \ref{fig:barplot_lines_micrornaseq_colorectal_cancer_res2_auc}. More 
  performance measures comparisons are presented in Section [! section ref here]
  of the SM. In addition, we investigated the sensitivity of
  the multiplier estimation using this data set in SM Section 
  [! section ref here]. A small sidenote: We experienced numerical issues with 
  \texttt{GRridge} when cross-validating 
  the global penalty parameter on the training data. We therefore estimated this
  global parameter on the full data for \texttt{GRridge}.

  The penalty multipliers estimated by 
  \texttt{gren} and \texttt{GRridge} are according to expectation: the highly
  expressed group receives the lowest penalty, followed by the medium expressed
  group, while the non-expressed group receives the strongest penalty. 
  \texttt{gren} seems to outperform the other methods for a range of model 
  sizes. This is especially pronounced for the larger models. The performance of
  \texttt{cMCP} is somewhat unstable. We noticed this is in general
  for the group-lasso and its extensions (see SM [! section ref here]). Overall,
  it seems as though the inclusion of the extra information on the features 
  benefited predictive performance: both \texttt{GRridge} and \texttt{gren} 
  outperform their respective non-group-regularized versions, albeit it only
  slightly for \texttt{GRridge}.
<<barplot_lines_micrornaseq_colorectal_cancer_res2_auc, fig.cap="Results for the colorectal cancer example with (a) the estimated penalty multipliers and (b) estimated AUC.", out.width="100%", fig.asp=1/2>>=
@

  \section{Application to RNAseq in lymph node cancer}\label{sec:applymph}
  In this application, the aim is to predict lymph node metastasis in patients 
  with HPV-negative oral cancer using sequenced RNAs from TCGA 
  \cite[]{the_cancer_genome_atlas_network_comprehensive_2015}. The features are 
  normalised TCGA RNASeqv2 profiles of head-and-neck squamous cell 
  carcinomas on 133 patients with HPV-negative tumours in the oral cavity. Of 
  these patients, 76 suffered from lymph node metastasis, while 57 did not. 
  For a thorough introduction of these data, see \cite{te_beest_improved_2017}.

  As co-data we considered two sources: (a) the cis-correlation between the
  the RNASeqv2 data and DNA copy numbers on the same patients, 
  quantified by Kendall's $\tau$ and binned into five groups, and (b) 
  $p$-values from an external microarray data set described in 
  \cite{mes_prognostic_2017}, again binned into five groups. We expect features 
  with a large positive Kendall's $\tau$ to be more important in 
  metastasis prediction \cite[]{masayesva_gene_2004}. Likewise, we expect 
  features with low $p$-values to be more important.

  We compared \texttt{gren} to the same methods as in Section 
  \ref{sec:appcolon}. However, the co-data consists of two partitions with 
  overlapping groups, so we use an extension of \texttt{cMCP} that is able to 
  handle multiple groupings \cite[]{zeng_overlapping_2016}. We split the data 
  into 120 training instances and 97 test instances, trained
  the models on the training set and estimated AUC on the test set. We present 
  the results in Figure 
  \ref{fig:barplot_lines_rnaseq_lymph_node_metastasis_res1_auc}. Other 
  comparisons and performance measures are presented in the SM Section 
  [! section ref here].

  The estimated penalty multipliers follow the expected pattern. Larger 
  cis-correlations receive smaller penalties. Likewise, smaller $p$-values are
  also penalized less. This pattern is slightly less pronounced in \texttt{gren}
  with $\alpha=0.95$ than in the other methods. The predictive performance of
  \texttt{gren} is slighty better than the other methods for a range of model
  sizes. We also note that the overlapping group \texttt{cMCP} is much more
  stable than the regular \texttt{cMCP} (see in SM [! section ref here]), 
  where we use as grouping the cross-tabulation of the two paritions. In this 
  example, ridge and \texttt{GRridge} perform almost identical.
<<barplot_lines_rnaseq_lymph_node_metastasis_res1_auc, fig.cap="Results for the lymph node metastasis example with (a) the estimated penalty multipliers for the cis-correlations, (b) the estimated penalty multipliers for the p-values, and (c) the estimated AUC. OcMCP refers to the overlapping group cMCP.", out.width="100%", fig.asp=1>>=
@

@@

	Although our method can be considered weakly adaptive, it is different from 
	the adaptive lasso \cite[]{van_de_geer_adaptive_2011,huang_adaptive_2008,
	zhang_adaptive_2007,zou_adaptive_2006} and adaptive elastic net 
	\cite[]{zou_adaptive_2009} in the sense that it adapts to external information
	rather than to the data. It also differs in the scale of adaptation: in the 
	adaptive lasso and elastic net the adaptive weights are feature specific,
	while in our case they are estimated on the group level, rendering the 
	adaptation more robust. As can be seen from both the simulations in Section 
	\ref{sec:simulations} and the application examples with different types of 
	outcome data (prognostic and diagnostic) and marker types (RNAseq, microRNAs, 
	and metabolomics) in Sections [! section refs here], and the SM
	Sections [! section refs here], adaptation to 
	external data may be beneficial for prediction and feature selection. We 
	believe that this is due to the `borrowing of information' effect: estimates 
	that are believed to be similar are shrunken in the same way, yielding 
	overall, better estimates.

@@

	\section*{Software}
	A (stable) \texttt{R} package is available from 
	\url{https://CRAN.R-project.org/package=gren}. All materials needed to   
	recreate this document are available from 
	\url{https://github.com/magnusmunch/gren}.

	\section*{Supplementary Material}
	Supplementary Material is available online from 
	\url{http://pub.math.leidenuniv.nl/~munchmm/}.

	\section*{Acknowledgements}
	\textit{Conflict of Interest}: None declared. We thank the editors and 
	referees for their useful suggestions and comments.

### 2019-03-01 docs/supplement.Rnw
ran micrornaseq_colorectal_cancer.R 10 random groups and stability selection

  \subsection{Random groups}
  Considering that the features may be partitioned into many groups, with one 
  parameter per group, we have to be aware of overfitting risks. We investigated
  this using the data introduced in Section [! section ref here] of the Main 
  Document, randomly dividing 
  the features into three groups. We fixed the group sizes to the group sizes 
  used in the three-group setting in Section [! section ref here] of the Main
  Docuemtn. Under this random 
  partitioning of the features, we expect all penalty multipliers to be 
  estimated as one if no overfitting occurs.

  We compared the estimated multipliers to the estimates by \texttt{GRridge}. 
  Since these results depend on one specific randomisation of the groups, we 
  repeated the procedure 100 times and present the results in [! figure ref]. 
  From this figure we see that the estimates for \texttt{gren} are close to one.
  The estimates by \texttt{GRridge} show much more variation. Additionally, 
  \texttt{GRridge} gives slightly biased estimates: the median penalty parameter
  estimates are [! medians here] for the three groups. In contrast, the median 
  estimates of \texttt{gren} are [! medians here].

  We repeated the simulation with ten evenly sized groups. Figure 
  [! figure ref here] presents the estimated penalty multipliers for 100 random 
  splits of the microRNA features into 10 evenly sized groups for 
  \texttt{GRridge} and \texttt{gren} with $\alpha \in \{ 0.05, 0.5, 0.95 \}$. 
  From the Figure we see that \texttt{GRridge} shows more variation around one 
  then the other models. In addition, the median penalty multipliers for the 
  three \texttt{gren} models are all between [! medians here], while the 
  \texttt{GRridge} median estimates range from [! medians here]. This indicates 
  that \texttt{GRridge} suffers from a slight bias in the estimation, while the 
  \texttt{gren} estimates appear to be unbiased.

@@

>>>>

  \subsection{Random groups}
  Considering that the features may be partitioned into many groups, with one 
  parameter per group, we have to be aware of overfitting risks. We investigated
  this using the data introduced in Section [! section ref here] of the MD, 
  randomly dividing the features into three groups. We fixed the group sizes to 
  the group sizes used in the MD. Under this random 
  partitioning of the features, we expect all penalty multipliers to be 
  estimated as one if no overfitting occurs.

<<>>=
res <- read.table("results/micrornaseq_colorectal_cancer_res3.csv", 
                  stringsAsFactors=FALSE)
medians <- round(apply(res, 2, median), 2)
@
  We compared the estimated multipliers to the estimates by \texttt{GRridge}. 
  Since these results depend on one specific randomisation of the groups, we 
  repeated the procedure 100 times and present the results in Figure 
  \ref{fig:boxplots_micrornaseq_colorectal_cancer_res3_multipliers}. 
  From this figure we see that the estimates for \texttt{gren} are close to one.
  The estimates by \texttt{GRridge} show much more variation. We found no strong
  indication of bias, with median estimated penalty parameters 
  \Sexpr{medians[c(1:3)]} for
  \texttt{GRridge} and \Sexpr{medians[c(4:6)]}; \Sexpr{medians[c(7:9)]},
  and \Sexpr{medians[c(10:12)]} for \texttt{gren} with 
  $\alpha \in \{ 0.05, 0.5, 0.95 \}$, respectively.
<<boxplots_micrornaseq_colorectal_cancer_res3_multipliers, fig.cap="Results for the colorectal cancer example with 100 realisations of 3 random groups of features.", out.width="50%", fig.asp=1>>=
@

  We repeated the simulation with ten evenly sized groups. Figure 
  [! figure ref here] presents the estimated penalty multipliers for 100 random 
  splits of the microRNA features into 10 evenly sized groups for 
  \texttt{GRridge} and \texttt{gren} with $\alpha \in \{ 0.05, 0.5, 0.95 \}$. 
  From the Figure we see that \texttt{GRridge} shows more variation around one 
  then the other models. In addition, the median penalty multipliers for the 
  three \texttt{gren} models are all between [! medians here], while the 
  \texttt{GRridge} median estimates range from [! medians here]. This indicates 
  that \texttt{GRridge} suffers from a slight bias in the estimation, while the 
  \texttt{gren} estimates appear to be unbiased.
<<boxplots_micrornaseq_colorectal_cancer_res4_multipliers, fig.cap="Results for the colorectal cancer example with 100 realisations of 10 random groups of features, plotted (a) with outliers, and (b) without outliers.", out.width="100%", fig.asp=1/2>>=
@

@@

  \section{Application to RNAseq in lymph node cancer}
  \texttt{cMCP} is not designed to deal with overlapping groups, so we 
  cross-tabulate the two co-data partitions to create one paritioning with 25 
  groups for \texttt{cMCP}.  
  Additional comparisons to \texttt{SGL}, \texttt{gel} and 
  the overlapping group version of \texttt{gel} \cite[]{zeng_overlapping_2016}
  \cite[]{breheny_group_2015} are found in the SM Section [! section ref here].
